Different forms of uncertainty differently affect the evolution of social learning

### 2022-07-29 ms.tex
edits to analysis and model

\begin{abstract}
Social learning is essential to human survival---it is a critical adaptation for dealing with uncertainty. However, uncertainty takes many forms. We identified four theoretically important types of uncertainty: temporal environmental variability, ambiguous payoffs, decision set size, and effective lifespans. We used this framework to study conditions for the evolution of social learning in populations with cognitively realistic individual learning capacities.
We used evolutionary agent-based modeling to track how often social learning evolved. Agents performed one of several behaviors to acquire payoffs. Individual learning used the softmax algorithm. Social learning occurs between generations  through vertical and oblique transmission.
Different types of uncertainty had different effects. Temporal environmental variability suppressed social learning. Larger decision set sizes promoted social learning. Payoff ambiguity and lifespan had more complex effects on social learning, dependent on other uncertainty parameters.
This shows the importance of clearly operationalizing uncertainty in theories of social learning evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

Empirically, the number of learning opportunities can be manipulated in the lab, but will also correspond with an individual's age the real world. In multi-round studies of information use in novel tasks, US participants' use of social information declined precipitously across rounds ~\cite{McElreath2005}, suggesting they were more likely to use social information when they were most uncertain about the task early on. In a more naturalistic context, Aplin et al. (2017) \nocite{Aplin2017} found that younger great tits more readily used social information compared with older individuals, possibly because they had accrued less information via individual learning, and also possibly because younger individuals have the most to gain (given their expected remaining lifespan) by switching to superior behavioral options. 
% \ps{I got rid of the Leris and Reader reference because it didn't seem relevant to age-based uncertainty. They just showed that young guppies exposed to reliable models were more likely to use social information later. I also added some human studies, which seems important.} 
Cross-cultural studies on humans have shown the importance of childhood as a phase of heavy social learning ~\cite{Reyes2016}. Young children are more likely to acquire their beliefs and simple skills from their parents than are older children or adults, which is at least partly due to the differential knowledge accrual between young children and the older adults to whom they direct the most trust and attention \cite{kline2013teaching}.
%The age of individuals was found to have an effect on social learning in both \emph{parus major}~\cite{Aplin2017} and \emph{poecilia reticulata} (guppies)~\cite{Leris2016}. \ps{What effect? I'm not sure }
%\mt{Need better connection between lifespan and age, or better references to support use of lifespan}.
Individuals with shorter effective lifespans will end their lives with more uncertainty about which behaviors are optimal, and thereby transmit some of that uncertainty to anyone who learns from them. The effective lifespan for learning varies across tasks, individuals, and species, yet most models assume one learning opportunity per generation ~\cite{Feldman1996,Henrich1998, perreault2012bayesian}. Some models do allow several cultural generations within 1 genetic generation ~\cite{Enquist2007}, but little formal theory explicitly examines the role of uncertainty based on number of learning opportunities into our understanding of the evolution of social learning.

@@

\subsection{Other adaptations to uncertainty}

@@

To understand how different forms of uncertainty interact to affect the evolution of
social learning, we measured how frequently social learning evolved across
uncertainty parameters. We analyze outcomes by plotting the mean observed social
learning frequency, $\meansl$, across all agents and 1000 simulation trials for each
uncertainty parameter setting.  We visualize and inspect plots with $\meansl$ on the
y-axis and environmental variability, $u$ on the x-axis. 
% This is convenient because we are most certain about how $\meansl$ depends on $u$:
% we expect $\meansl$ to decrease as $u$ increases since environmental variability
% is well understood to cause socially-transmitted information to become outdated. 
To inspect patterns in $\meansl$ over all uncertainty parameters, we analyze nine
plots of $\meansl$ over $u$ in a $3\times3$ plot array
(Figure~\ref{fig:mainResults}).  Plot array columns correspond to different
selection set sizes, $B$; rows correspond to different ambiguous payoffs via
non-optimal expected payoffs, $\pilow$ (recall $\pihigh=0.9$ for all analyses);
finally, plot line and marker colors correspond to different effective lifespans,
$L$.  
% We support our conclusions regarding the evolution of social learning by analyzing
% the time to fixation and by comparing the average net payoffs in our simulations
% to the expected payoffs for individuals in both populations of all social learners
% and in populations of all individual learners. We thus use our analysis to
% understand changes to how $\meansl$ changes over $u$ for different uncertainty
% parameter settings.

Several patterns emerged in changes to social learning prevalence over
the uncertainty parameters (Figure~\ref{fig:mainResults}). 
First, we indeed observed that greater $u$ uniformly decreased
$\meansl$, across all other uncertainty parameter values. The other uncertainty
parameters changed when $\meansl$ began to decrease in terms of $u$,
and how quickly $\meansl$ decreased from 1 to 0 over $u$. For example, 
when payoff ambiguity and selection set size were
both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when $u$ was
small.
This is a relatively easy problem, and individual learning is more
efficient in this case (Figure~\ref{fig:mainResults}, upper left).
As $\pilow$ increased, $\meansl$ decreased less rapdily over $u$, 
which was more pronounced for smaller $L$ (see rows of Figure~\ref{fig:mainResults}).

Increased selection set size $B$ led nearly uniformly to more prevalent social
learning with all other uncertainty parameters being equal ($B$ increases along
columns in Figure~\ref{fig:mainResults}).  This is because misleading social
information is easy to overcome for two reasons: (1) larger $B$ presents a more
difficult problem for trial-and-error learning; and (2) a misidentified optimal
behavior is weighted less relative to smaller $B$, meaning agents are more likely to
explore behaviors identified as non-optimal when $B$ is larger (columns of
Figure~\ref{fig:mainResults}).  
To understand point (2), consider the case where a social learner child learns
that behavior 1 has an expected payoff of 0.8, and the rest have an expected
payoff of 0, but the optimal behavior changes between generations. Assume
the child tries behavior 1, but it pays off 0.
If $B=2$, then at within-generation $t=2$, $\ledger = (0.4, 0.0)$. The probability
the agent explores behavior 2 is now 0.02. However, if $B=10$ the agent has
$\ledger = (0.4, 0.0, \ldots, 0.0)$; the probability
the agent explores one of the nine alternative behaviors increases to 0.14.
If the agent again tries behavior 1, the probability of exploring alternatives
with $B=2$ is 0.21, but with $B=10$ the probability of exploring alternatives
greatly increases to 0.71. This explanation does not apply when $L=1$, where
social learning is increasingly beneficial as $B$ increases because it is more 
valuable to be possibly misled by outdated information than it is for agents
to blindly guess which is optimal.

@@

As a final note, we believe 
our modeling approach may have applications to developing social artificial
intelligence, sepcifically multi-agent reinforcement learning
systems~\cite{Sandholm1996,Ndousse2021,Gronauer2022,Jaques2019}.
It seems to be implicitly assumed by current multi-agent reinforcement learning
algorithms that social learning
would always be beneficial. However, we have shown in our work that sometimes 
it is more efficient for autonomous social agents to solve problems without
reference to social information across a variety of uncertainty conditions. 
Our model agents are essentially software
robots, so our results should predict the benefit to real-world social robots
to either learn socially or asocially if the assumptions we make here are 
reasonably met or implemented in the social robot design.
Our model dynamics could thus be used as an
evolutionary algorithm to update AI agents' social learning strategy without
explicit knowledge of, or need to approximate, the underlying uncertainty structure.
A designer would need only to program the robots to go through evolutionary
cycles of simulated ``reproduction'' and ``death''.
Our model could also predict the expected benefit of social learning, which may be
inherently costly in artificial multi-agent learning systems due to extra
energy or time that may be required to transmit information between agents.

>>>>

\begin{abstract} Social learning is essential to human survival---it is a critical
  adaptation for dealing with uncertainty. However, uncertainty takes many forms.
  We identified four theoretically important types of uncertainty: temporal
  environmental variability, ambiguous payoffs, decision set size, and effective
  lifespans. We used this framework to study conditions for the evolution of
  social learning in populations with cognitively realistic individual learning
  capacities.  We used evolutionary agent-based modeling to track how often social
  learning evolved. Agents performed one of several behaviors to acquire payoffs.
  Individual learning used the softmax algorithm. Social learning occurs between
  generations  through vertical and oblique transmission.  Different types of
  uncertainty had different effects. Temporal environmental variability suppressed
  social learning. Larger decision set sizes promoted social learning even
  when environmental change was more likely than not. Payoff
  ambiguity and lifespan had more complex effects on social learning, dependent on
  other uncertainty parameters.  This shows the importance of clearly
  operationalizing uncertainty in theories of social learning evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
\footnote{This document contains
\total{citnum} references.}  
\end{abstract}

@@

Empirically, the number of learning opportunities can be manipulated in the lab,
but will also correspond with an individual's age the real world. In multi-round
studies of information use in novel tasks, US participants' use of social
information declined precipitously across rounds ~\cite{McElreath2005}, suggesting
they were more likely to use social information when they were most uncertain
about the task early on. In a more naturalistic context, Aplin et al. (2017)
\nocite{Aplin2017} found that younger great tits more readily used social
information compared with older individuals, possibly because they had accrued
less information via individual learning, and also possibly because younger
individuals have the most to gain (given their expected remaining lifespan) by
switching to superior behavioral options. 
% \ps{I got rid of the Leris and Reader reference because it didn't seem relevant
% to age-based uncertainty. They just showed that young guppies exposed to
% reliable models were more likely to use social information later. I also added
% some human studies, which seems important.} 
Cross-cultural studies on humans have shown the importance of childhood as a phase
of heavy social learning ~\cite{Reyes2016}. Young children are more likely to
acquire their beliefs and simple skills from their parents than are older children
or adults, which is at least partly due to the differential knowledge accrual
between young children and the older adults to whom they direct the most trust and
attention \cite{kline2013teaching}.
%The age of individuals was found to have an effect on social learning in both
%\emph{parus major}~\cite{Aplin2017} and \emph{poecilia reticulata}
%(guppies)~\cite{Leris2016}. \ps{What effect? I'm not sure } \mt{Need better
%connection between lifespan and age, or better references to support use of
%lifespan}.
Individuals with shorter effective lifespans will end their lives with more
uncertainty about which behaviors are optimal, and thereby transmit some of that
uncertainty to anyone who learns from them. The effective lifespan for learning
varies across tasks, individuals, and species, yet most models assume one learning
opportunity per generation ~\cite{Feldman1996,Henrich1998, perreault2012bayesian}.
Some models do allow several cultural generations within 1 genetic generation
~\cite{Enquist2007}, but little formal theory explicitly examines the role of
uncertainty based on number of learning opportunities into our understanding of
the evolution of social learning.

@@

% \subsection{Other adaptations to uncertainty} 
\subsection{Individual-level adaptations to uncertainty}

@@

To understand how different forms of uncertainty interact to affect the evolution of
social learning, we measured how frequently social learning evolved across
uncertainty parameters. We analyze outcomes by plotting the mean observed social
learning frequency, $\meansl$, across all agents and 1000 simulation trials for each
uncertainty parameter setting.   
% This is convenient because we are most certain about how $\meansl$ depends on $u$:
% we expect $\meansl$ to decrease as $u$ increases since environmental variability
% is well understood to cause socially-transmitted information to become outdated. 
To inspect patterns in $\meansl$ over all uncertainty parameters, we analyze nine
plots of $\meansl$ over $u$ in a $3\times3$ plot array
(Figure~\ref{fig:mainResults}). All plots show how $\meansl$ changes (y-axis)
with environmental variability, $u$ (x-axis). Selection set size, $B$ increases
from left to right in the plot array (columns). Payoff ambiguity via $\pilow$ 
increases from top to bottom in the array (recall $\pihigh=0.9$ for all analyses).
Effective lifespan, $L$, is indicated by plot line and marker colors. 
Several patterns in social learning prevalence emerged as 
the uncertainty parameters were varied (Figure~\ref{fig:mainResults}). 
% We support our conclusions regarding the evolution of social learning by analyzing
% the time to fixation and by comparing the average net payoffs in our simulations
% to the expected payoffs for individuals in both populations of all social learners
% and in populations of all individual learners. We thus use our analysis to
% understand changes to how $\meansl$ changes over $u$ for different uncertainty
% parameter settings.

First, we observed that greater $u$ uniformly decreased
$\meansl$ across all other uncertainty parameter values. The other uncertainty
parameters changed at which $u$ $\meansl$ began to decrease,
and how rapidly $\meansl$ decreased over $u$. For example, 
when payoff ambiguity and selection set size were
both small ($\pilow = 0.1$ and $B=2$), social learning evolved only when $u$ was
small. These parameter settings present a relatively easy problem for individuals.
Furthermore, it is relatively costly for agents to be misled by outdated social 
information in this case, especially for longer lifespans for finding the 
optimal behavior (Figure~\ref{fig:mainResults}, upper left). 
In another example, when $\pilow = 0.8$, $\meansl$ decreased less rapdily over $u$ due to 
weak selection caused by little advantage for performing the optimal versus
non-optimal behavior (see rows of Figure~\ref{fig:mainResults}).

Increased selection set size $B$ led nearly uniformly to more prevalent social
learning, all other uncertainty parameters static. Environments with 
larger $B$ present a more difficult problem for trial-and-error learning within
a lifespan. This effect is amplified because a misidentified optimal
behavior is easier to correct for larger $B$ relative to smaller $B$.  
To understand this, consider the case where a social learner child learns
that behavior 1 has an expected payoff of 0.8, and the rest have an expected
payoff of 0, but the optimal behavior changes between generations. Assume
the child tries behavior 1, but it pays off 0.
If $B=2$, then at within-generation $t=2$, $\ledger = (0.4, 0.0)$. The probability
the agent explores behavior 2 is now 0.02. However, if $B=10$ the agent has
$\ledger = (0.4, 0.0, \ldots, 0.0)$; the probability
the agent explores one of the nine alternative behaviors increases to 0.14.
If the agent again tries behavior 1, the probability of exploring alternatives
with $B=2$ is 0.21, but with $B=10$ the probability of exploring alternatives
greatly increases to 0.71. This explanation does not apply when $L=1$, where
social learning is increasingly beneficial as $B$ increases because it is more 
valuable to be possibly misled by outdated information than it is for agents
to blindly guess which is optimal.

@@

% As a final note, we believe 
% our modeling approach may have applications to developing social artificial
% intelligence, sepcifically multi-agent reinforcement learning
% systems~\cite{Sandholm1996,Ndousse2021,Gronauer2022,Jaques2019}.
% It seems to be implicitly assumed by current multi-agent reinforcement learning
% algorithms that social learning
% would always be beneficial. However, we have shown in our work that sometimes 
% it is more efficient for autonomous social agents to solve problems without
% reference to social information across a variety of uncertainty conditions. 
% Our model agents are essentially software
% robots, so our results should predict the benefit to real-world social robots
% to either learn socially or asocially if the assumptions we make here are 
% reasonably met or implemented in the social robot design.
% Our model dynamics could thus be used as an
% evolutionary algorithm to update AI agents' social learning strategy without
% explicit knowledge of, or need to approximate, the underlying uncertainty structure.
% A designer would need only to program the robots to go through evolutionary
% cycles of simulated ``reproduction'' and ``death''.
% Our model could also predict the expected benefit of social learning, which may be
% inherently costly in artificial multi-agent learning systems due to extra
% energy or time that may be required to transmit information between agents.
